---
title: 'STAT 413/613 HW 2: stringr and lubridate'
author: "Chi Yun Liu"
date: "`r Sys.Date()`"
output:
  html_document: 
    number_sections: true
editor_options:
  chunk_output_type: inline
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align  = "center",
                      fig.height = 3, 
                      fig.width  = 5)
```

```{r include=FALSE}
library(tidyverse)
#(0213:- 0 No need to load ggplot after tidyverse)
#library(ggplot2)
```


# Instructions {-}

Rename the starter file under the analysis directory as `hw_02_yourname.Rmd` and use it for your solutions.   
1. Modify the "author" field in the YAML header.  
2. Stage and Commit R Markdown and HTML files (no PDF files).   
3. **Push both .Rmd and HTML files to GitHub**.   
- Make sure you have knitted to HTML prior to staging, committing, and pushing your final submission.  
4. **Commit each time you answer a part of question, e.g. 1.1**   
5. **Push to GitHub after each major question**, e.g., Scrabble and Civil War Battles   
- **Committing and Pushing are graded elements for this homework.**  
6. When complete, submit a response in Canvas  
    
- Only include necessary code to answer the questions.
- Most of the functions you use should be from the tidyverse. Too much base R will result in point deductions.
- Use Pull requests and or email to ask me any questions. If you email, please ensure your most recent code is pushed to GitHub.  

- Learning objectives:
    - Manipulate dates and times with lubridate.


# Scrabble Words

For this exercise, we are using the [Collins Scrabble Words](https://en.wikipedia.org/wiki/Collins_Scrabble_Words), 
which is most commonly used outside of the United States. The dictionary most often used in the United States is the 
[Tournament Word List](https://en.wikipedia.org/wiki/Official_Tournament_and_Club_Word_List).

**WARNING**: Do not try `str_view()` or `str_view_all()` on these data.It will stall your computer.

1. Use a readr function to load the 2015 list of Collins Scrabble Words into R from your data folder or from 
   <https://data-science-master.github.io/lectures/data/words.txt>  
   + (note: "NA" is an official Scrabble word).
```{r}
scwords <- read_csv(file = "../data/words.txt")
head(scwords) 
```
   
   
2. What are the six longest words that have the most "X"'s in them?

```{r}
scwords_df <- as.data.frame(scwords)
head(scwords_df)

scwords_df[is.na(scwords_df)] <- "NA"

scwords_df %>% 
  mutate(detect_x = str_count(word, "[Xx]")) %>% 
  arrange(desc(detect_x)) %>% 
  mutate(word_long = str_length(word)) %>%
  filter(detect_x == 2) %>% 
  arrange(desc(word_long)) %>% 
  slice_head(n = 6)

# COEXECUTRIXES, EXTRATEXTUAL, COEXECUTRIX, EXECUTRIXES, SAXITOXINS, XANTHOXYLS
```


3. How many words have an identical first and second half of the word? If a word has an odd number of letters, exclude the middle character.  
  - MURMUR counts because MUR is both the first and second half.
  - JIGAJIG counts because the middle A is excluded so JIG is both the first and second half. 
  - Save the results to a variable.
  
```{r}
scwords_df %>% 
  mutate(word_count = str_length(word)%%2) %>% 
  filter(word_count == 0) %>% 
  mutate(word_even = str_length(word)) %>%
  mutate(even_mid = word_even/2) %>% 
  mutate(first_even = str_sub(word, 1, even_mid)) %>% 
  mutate(sec_even = str_sub(word, even_mid+1, word_even)) %>% 
  filter(first_even == sec_even) -> even_sum
head(even_sum)

scwords_df %>% 
  mutate(word_count = str_length(word)%%2) %>% 
  filter(word_count == 1) %>% 
  mutate(word_odd = str_length(word)) %>% 
  mutate(odd_mid = (word_odd + 1)/2) %>% 
  mutate(first_odd = str_sub(word, 1, odd_mid-1)) %>% 
  mutate(sec_odd = str_sub(word, odd_mid+1, word_odd)) %>% 
  filter(first_odd == sec_odd) -> odd_sum
head(odd_sum)

q3_ans <- count(even_sum) + count(odd_sum)
q3_ans
# 254 words have an identical first and second half of the word.
```
  

4. Use the results from 3 to find the longest word with an identical first and second half of the word?

```{r}
even_sum %>% 
  arrange(desc(word_even)) %>% 
  slice_head(n=1)

odd_sum %>% 
  arrange(desc(word_odd)) %>% 
  slice_head(n=1)

# CHIQUICHIQUI is the longest word with an identical first and second half of the word
```


# Civil War Battles

The data in "civil_war_theater.csv" contains a information on American Civil War battles, taken from [Wikipedia](https://en.wikipedia.org/wiki/List_of_American_Civil_War_battles).  

Variables include:   

- `Battle`: The name of the battle.
- `Date`: The date(s) of the battle in different formats depending upon the length of the battle. 
  + If it took place on one day, the format is "month day, year". 
  + If it took place over multiple days, the format is "month day_start-day_end, year". 
  + If it took place over multiple days and months, the format is "month_start day_start - month_end day_end, year". 
  + If it took place over multiple days,months, and years, the format is "month_start day_start, year_start - month_end day_end, year_end".
- `State`: The state where the battle took place. Annotations (e.g.     describing that the state was a territory at the time) are in parentheses.
- `CWSAC`: A rating of the military significance of the battle by the Civil War Sites Advisory Commission. `A` = Decisive, `B` = Major, `C` = Formative, `D` = Limited.
- `Outcome`: Usually `"Confederate victory"`, `"Union victory"`, or `"Inconclusive"`, followed by notes.
- `Theater`: An attempt to to identify which theater of war is most associated with the battle
    
1. Use a readr function and relative paths to load the data into R.

```{r}
war_df <- read_csv("../data/civil_war_theater.csv")
head(war_df)
```
    
The next several questions will help you take the dates from all the different formats and create a consistent set of start date and end date variables in the data frame. We will start by calculating how many years, and months are in each battle.

2. Add a variable to the data frame with the number of years for each battle.  

- Create a character variable as follows. This can be used as a pattern in a regular expression.  
   
- Use `year_regex` to now count the number of years in each battle, add this to the data frame,  and save the data frame.

```{r}
year_regex <- stringr::str_c(1861:1865, collapse = "|")
year_regex


war_df %>% 
  mutate(num_years = str_count(Date, year_regex)) %>% 
  select(Battle, Date, num_years, everything())->war_df
head(war_df)

```

3. Add a variable to the data frame with the number of months for each battle.  
- Consider R's built-in vector of month names: `month.name`.
- Use `month.name` to count the number of month names in the `Date` variable in each battle. 
- Add this to the data frame. (You might need to do something similar to what we did in part 2).

```{r}
month_regex <- stringr::str_c(month.name, collapse = "|")
month_regex

war_df %>% 
  mutate(num_month = str_count(Date, month_regex)) %>%
  select(Battle, Date, num_month, everything())->war_df
head(war_df)
```

4. Add a variable to the data frame that is `TRUE` if `Date` spans multiple days and is `FALSE` otherwise. Spanning multiple months and/or years also counts as `TRUE`.

```{r}
war_df %>% 
  mutate(dmulti_span = str_detect(Date, "-")) %>% 
  select(Battle, Date,dmulti_span, num_month, everything())->war_df
head(war_df)
```

   
5. Make four new data frames by filtering the data based on the length of the battles:  
   + a data frame with the data for only those battles spanning just one day, 
```{r}
war_df %>% 
  filter(dmulti_span == FALSE) -> one_day
head(one_day)
```
   
   + a data frame with the data for only those battles spanning multiple days in just one month,
```{r}
war_df %>% 
  filter(num_month == 1 & dmulti_span == TRUE) ->one_mon
head(one_mon)
```
   
   + a data frame with the data for only those battles spanning multiple months but not multiple years, and,
```{r}
war_df %>% 
  filter(num_month > 1 & num_years <2)->multi_mon
head(multi_mon)
```
   
   + a data frame with the data for only those battles spanning multiple years.
```{r}
war_df %>% 
  filter(num_years > 1)->multi_years
head(multi_years)
```
 
6. For each of the four new data frames,  
  - Add two new variables:  
    + `Start` should contain the start-date of each battle. 
    + `End` should contain the end-date of each battle.
    + Hint: look at help for `separate()`
    + Make sure these are `Date` class objects. 
  - Remove the `Date` variable from  each data frame.
  - Save the data frames with the new variables
  
```{r}
one_day %>% 
  mutate(Start = parse_date(Date, format = "%B %d, %Y")) %>% 
  mutate(End = parse_date(Date, format = "%B %d, %Y")) %>% 
  select(Battle, Start, End, everything(), -Date)->
  one_day
head(one_day)
#(0213:2.2.61:  - 0 Easier to use lubridate than readr)
```

```{r}
one_mon %>% 
  separate(Date, into = c("Month", "Year"), sep = ",") %>% 
  separate(Month, into = c("Month", "Sday"), sep = " ") %>% 
  separate(Sday, into = c("Sday", "Eday"), sep = "-") %>% 
  mutate(Day_span = as.double(Eday) - as.double(Sday)) %>% 
  unite(Start, Month, Sday, Year, sep = "-") %>% 
  mutate(Start = parse_date(Start, format = "%B-%d- %Y")) %>% 
  mutate(End = Start + Day_span) %>% 
  select(Battle,Start, End, everything(), -Eday, -Day_span)-> one_mon
head(one_mon)
```

  
```{r}
multi_mon %>%
  separate(Date, into = c("dspan", "Year"), sep = ",") %>% 
  separate(dspan, into = c("Sdate", "Edate"), sep = "-" ) %>% 
  mutate(new_sdate = str_trim(Sdate)) %>% 
  mutate(year_1 = Year) %>% 
  unite(new_sdate, Year, col = "SSdate", sep = ",") %>% 
  unite(Edate, year_1, col = "EEdate", sep = ",") %>% 
  mutate(Start = parse_date(SSdate, format = "%B %d, %Y")) %>% 
  mutate(End = parse_date(EEdate, format = "%B %d, %Y")) %>% 
  select(Battle, Start, End, everything(), -Sdate, -EEdate, -SSdate,)->multi_mon
head(multi_mon)
```
  
```{r}
multi_years %>% 
  separate(Date, into = c("Start", "End"), sep = "-") %>%
  mutate(new_start = str_trim(Start)) %>% 
  mutate(Start = parse_date(new_start, format = "%B %d, %Y")) %>% 
  mutate(End = parse_date(End, format = "%B %d, %Y")) %>% 
  select(Battle, Start, End, everything(), -new_start)->multi_years
head(multi_years)
```

7. Create a new data frame with all the battles and the Start and End  dates by binding the rows of the four data frames as updated in part 6 

```{r}
civil_war <- bind_rows(list(one_day, one_mon, multi_mon, multi_years))
head(civil_war)
```

8. Calculate the number of days each battle spanned.  
- What's the longest battle of the war? 
- How long did it last?

```{r}
civil_war %>% 
  mutate(Span = End - Start) %>% 
  select(Battle, Start, End, Span, everything()) %>% 
  arrange(desc(Span)) %>% 
  slice_head(n = 1) 
# Siege of Port Hudson is the longest battle, and it last 49 days long.
```

9. Is there an association between the CWSAC significance of a battle and its duration?  
- Create an appropriate plot. 
- Interpret the plot in one sentence to answer the question.
- Extra Credit: Test for a linear relationship using `lm()` and interpret the results in one sentence based on the $p$-value and adjusted R-squared.

```{r}
war_dur <- civil_war$End - civil_war$Start

civil_war %>% 
  ggplot(mapping = aes(x = CWSAC, y = (war_dur))) +
  geom_boxplot() +
  labs(title = "The CWSAC significance of a battle versus its duration",
       y = "duration")

# The boxplot shows that the level of A is more spread in duration with higher median, but overall CWSAC significance of a battle are generally similar in its duration that spreading around 0-3 days.
```

```{r}
lm_model <- lm(as.double(war_dur) ~ as.factor(CWSAC), data = civil_war)
summary(lm_model)
# The model is significant since the p-value is extremely small(2.026e-05<.05), but only 5.48% of the variability is explained by the model. 

# (0213): - 0.1 The skewness of the data suggests a transformation to make it more symmetrical
```

10. Extra Credit: Did the [theaters of war](https://en.wikipedia.org/wiki/Theater_(warfare)) shift during the American Civil War?  
- Reproduce this plot in R (filter out states with two or fewer battles.)
- Interpret the plot in one sentence.

```{r}
# Combine state with the same name (rename input)
# filter n > 2
# deal with plot: state names
# plot

# rename input
civil_war %>% 
  mutate(State = str_replace_all(State, "West Virginia \\(Virginia at the time\\)", "West Virginia")) %>% 
  mutate(State = str_replace_all(State, "North Dakota \\(Dakota Territory at the time\\)", "North Dakota")) %>% 
  mutate(State = str_replace_all(State, "North Dakota \\(Dakota Territory..............", "North Dakota")) %>% 
  mutate(State = str_replace_all(State, "Oklahoma \\(Indian Territory at the time\\)", "Oklahoma")) -> civil_war01

unique(civil_war01$State)

#civil_war01 %>% 
  #filter(State == "West Virginia")

# filter n <= 2
civil_war01 %>%
  group_by(State) %>%
  count() %>%
  filter(n <= 2) -> civil_war02

civil_war01 %>% 
  anti_join(civil_war02, by = "State") %>% 
  mutate(Theater_new = Theater) %>% 
  unite(State, Theater_new, col = "state_thr", sep = "-") %>% 
  select(Battle, state_thr, Start, Theater) -> civil_war01

civil_war01 %>% 
  #filter(Theater == "Lower Seaboard" | Theater == "Western") %>%
  group_by(state_thr) %>% 
  #count() %>% 
  mutate(state_thr = str_replace_all(state_thr, "Lower Seaboard", "Lo")) %>% 
  mutate(state_thr = str_replace_all(state_thr, "Western", "We")) %>% 
  mutate(state_thr = str_replace_all(state_thr, "\\-Eastern", "")) %>% 
  mutate(state_thr = str_replace_all(state_thr, "\\-Trans-Mississippi", "")) %>% 
  mutate(state_thr = as.factor(state_thr))->civil_war01

unique(civil_war01$state_thr)

civil_war01 %>%
  ggplot(mapping = aes(x = Start, y = state_thr, fill =Theater)) +
  geom_boxplot()+
  labs(x = "Start", y = "State")

#(0213)- 0.5 Needed to combine  states names with Theater for those states into two theaters and then fct_reorder2(StateT, Start, Theater)
```

* The trend of purple bar in the plot shows that the theater did shift to the western and most of the theater are also in the western.

















